Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-20 paper/manuscript.tex
Lots of fiddling about with tables and checklist

computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\texbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 
Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}.

@@

\begin{equation}
    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}}{\int_\mathrm{unbound} \exp(-\beta U(\vec{q}))d\vec{q}},
\end{equation}
where $\beta = \frac{1}{k_BT}$ and $F$ is the Helmholtz free energy. 
Or more generally in a simplified form the free energy difference ($\Delta f$) between the bound and unbound states is:
\begin{equation}
\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
\end{equation}
where $Z = \int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}$ represents the configurational integral. 
Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 
So, how can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation? Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation, as typically binding events happen at a $\mu$s scale, and simulations reach around 1 $\mu$s/day. For a reliable estimate of a single free energy of binding many days of simulations would be required, making these type of calculations uninteresting at a pharamaceutical timescale. Instead, we can run alchemical free energy simulations, by making use of the fact that free energy is a state function. We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation. Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one. Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B). 
The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, e^{-u(\vec{q}; \lambda)} ,
\label{equation:partition-function-definition}
\end{eqnarray}
, where $\Gamma$ signifies the phase space volume over which to integrate. 
The \emph{reduced potential} $u(\vec{q}; \lambda)$~\cite{shirts2008statisticallya} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(\vec{q};\lambda) &\equiv& \beta \left[ U(\vec{q};\lambda) + p \, V(\vec{q}) + \sum_{i=1}^N \mu_i \, N_i(\vec{q}) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity, see for example the thermodynamic cycle for an absolute free energy calculation as seen in~\ref{fig:fig_absolute_thermodynamic_cycle}.

@@

Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently. Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported. \cite{REF Reproducibility of hydration free energies Loefler et al.  JCTC 2018} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{Ref Rizzi et al SAMPLing biorxiv} and, in many cases, single-point energies were identical or nearly so. Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions. 

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspectiv}. From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations. 

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene (A) to benzyl alcohol (B) Fig.~\ref{fig:topology}.

@@

\begin{figure}[h!]

@@

    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology} \textbf{left}: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. \textbf{right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.

@@

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemical}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.

@@

However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalized} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.

@@

Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically.
~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.

@@

With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common, different approaches have been designed to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed replicas each representing a different $\lambda$ value that run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A}). This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for Siremol/Openmm~\cite{}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighboring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B}). It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
\end{figure} 
The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.

@@

Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn. These additional observables include, but are not limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig. \ref{fig:convergence_forward_reverse}).
\begin{figure}[h!]

@@

For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions. [e.g. get ref from Sukanya]

@@

    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy \textbf{B}. A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue). Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}). \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.

@@

One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespace,wu2005phasespacea}. For this purpose, a so-called overlap matrix can be used. This is a $K\times K$ matrix, where $K$ is the number of simulated states, i.e. $\lambda$s. Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI. 

@@

In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference. Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller). In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e. there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae. Consequently, the threshold has to be derived empirically. It has been proposed that the values of the first off-diagonals (i.e. the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}. Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(C)), as the variance tends to be underestimated in case of poor overlap.

\begin{figure}
\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data. (A) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (B) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values. (C)shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
\end{figure}

Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (A), (B), and (C) respectively. For Fig.~\ref{fig:overlap}(A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good. In the case of Fig.~\ref{fig:overlap}(B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained. For Fig.~\ref{fig:overlap}(C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states. While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.

@@

Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section~\ref{}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

@@

Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g. adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking. Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways. This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence. Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.

@@

\todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}

@@

An example of a best practice scatter comparison between computed and experimental values is shown in~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals. The data for this plot is artificially generated for illustration purposes.

>>>>

computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 
Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

\item Complex association stoichiometry (beyond simple 1:1 association).

@@

\begin{equation}
    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}}{\int_\mathrm{unbound} \exp(-\beta U(\vec{q}))d\vec{q}},
\end{equation}
where $\beta = \frac{1}{k_BT}$ and $F$ is the Helmholtz free energy. 
Or more generally in a simplified form the free energy difference ($\Delta f$) between the bound and unbound states is:
\begin{equation}
\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
\end{equation}
where $Z = \int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}$ represents the configurational integral. 
Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 
So, how can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation? Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation, as typically binding events happen at a $\mu$s scale, and simulations reach around 1 $\mu$s/day. For a reliable estimate of a single free energy of binding many days of simulations would be required, making these type of calculations uninteresting at a pharamaceutical timescale. Instead, we can run alchemical free energy simulations, by making use of the fact that free energy is a state function. We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation. Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one. Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B). 
The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, \exp({-u(\vec{q}; \lambda)}) ,
\label{equation:partition-function-definition}
\end{eqnarray}
, where $\Gamma$ signifies the phase space volume over which to integrate. 
The \emph{reduced potential} $u(\vec{q}; \lambda)$~\cite{shirts2008statisticallya} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(\vec{q};\lambda) &\equiv& \beta \left[ U(\vec{q};\lambda) + p \, V(\vec{q}) + \sum_{i=1}^N \mu_i \, N_i(\vec{q}) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity, see for example the thermodynamic cycle for an absolute free energy calculation as seen in~\ref{fig:fig_absolute_thermodynamic_cycle}.

@@

Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently. Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported. \cite{loeffler2018reproducibility} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions. 

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations. 

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.

@@

@@

    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. \textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} (\textbf{B})), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.

@@

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemicala}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculationa}.

@@

However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalizeda} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.

@@

Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.

@@

With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common, different approaches have been designed to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed replicas each representing a different $\lambda$ value that run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A}). This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighboring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B}). It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{D}). 
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.

@@

Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn. These additional observables include, but are not limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig. ~\ref{fig:convergence_forward_reverse}).

@@

For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions~\cite{lincoff2016comparing}.

@@

    (\textbf{A}) The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy (\textbf{B}). A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue). Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}). (\textbf{E}) The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.

@@

One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}. For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.  $\mathcal{O} \in \mathbb{R}^{K\cross K}$, with $K$ being the number of simulated states, i.e. $\lambda$s. Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI. 

@@

In an overlap matrix $\mathcal{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference. Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller). In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e. there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae. Consequently, the threshold has to be derived empirically. It has been proposed that the values of the first off-diagonals (i.e. the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}. Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(\textbf{C})), as the variance tends to be underestimated in case of poor overlap.
\begin{figure}[ht!]
\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data. (\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values. (\textbf{C})shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (\textbf{A}), (\textbf{B}), and (\textbf{C}) respectively. For Fig.~\ref{fig:overlap}(\textbf{A}), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good. In the case of Fig.~\ref{fig:overlap}(\textbf{B}), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained. For Fig.~\ref{fig:overlap}(\textbf{C}), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states. While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.

@@

Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section~\ref{sec:simulation_protocol_choice}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

@@

Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g. adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking. Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019optimal,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways. This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence. Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.

@@

@@

An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals. The data for this plot is artificially generated for illustration purposes.
